---
title: "Michelle's research"
output: html_document
---

```{r}
library(tidyverse)
liver_data<-read.csv("ALL DATAlivercancer_Sort.csv")
head(liver_data,5)
tail(liver_data,5)
str(liver_data)
```

```{r}
dfu = lapply(liver_data[], unique)
dfu
```
#Drop trt, patient_workheet2, and id as those aren't what we're interested in. Drop HBV_PCR as it only has values 0 and NA, therfore probably not useful.

```{r}
naP = sum(is.na(liver_data$Platelets))
naP
naALT = sum(is.na(liver_data$ALT__SGPT_))
naALT
naAlK = sum(is.na(liver_data$Alk__Ph))
naAlK
naHCV = sum(is.na(liver_data$HCV_Ab))
naHCV
nahg = sum(is.na(liver_data$HG))
nahg
nab = sum(is.na(liver_data$bilirubin))
nab
naalb = sum(is.na(liver_data$S_Albumin))
naalb
naaAST = sum(is.na(liver_data$AST__SGOT_))
```
#Remove ALK__PH from model as more than half the observations have na values, therefore data wouldn't be that useful
#Remove or impute observations which have na values in Platelets, ALT__SGPT_, HCV_Ab, and HG to clean dataset. Remove for now.

```{r}
dataclean = na.omit(subset(liver_data, select = -c(trt, patient_Worksheet_2, id, HBV_PCR, Alk__Ph), data = liver_data))
str(dataclean)
```

```{r}
dfcu = lapply(dataclean[], unique)
dfcu
```
#Convert int-factors to type factor
```{r}
c = c("APC","FHIT","P15","P73","P14","P16","DAPK","RAR_","RASSF1A","O6MGMT","E_cadherin","hbs_ag1","methyl")
dataclean[c] <- lapply(dataclean[c], factor)
```
```{r}
str(dataclean)
```
#Full log model
```{r}
fit<-glm(type ~ ., data = dataclean, family = binomial())
```
```{r}
summary(fit)
```
#Check factor variables in regards to type to see if prob  = 1 for some factors

```{r}
xa = aggregate(x = liver_data[c], by = liver_data["type"], FUN = sum)
xa
```
#drop hbs_ag1 as only t patients have it
```{r}
fit2<-glm(type ~ .- hbs_ag1, data = dataclean, family = binomial())
summary(fit2)
```
```{r}
fit2b<-glm(type ~ APC + FHIT + P15 + P73 + P14 + P16 + DAPK + RAR_ + RASSF1A + O6MGMT  + E_cadherin + methyl,data = dataclean, family = binomial())
summary(fit2b) 
```
#Most significant factors are APC1 P151 and P141
#some kind of problem exists with the other non factors, check non factors 

```{r}
c2 = c("age", "gender", "WBCs", "HG","Platelets","AST__SGOT_","ALT__SGPT_","S_Albumin", "bilirubin") 
xa2 = aggregate(x = liver_data[c2], by = liver_data["type"], FUN = mean)
xa2
```
# remove wbc and Platelets as they cause fitted prob to be 1, not sure why
```{r}
fit3<-glm(type ~ age + HG + AST__SGOT_ + ALT__SGPT_ + S_Albumin + bilirubin, data = dataclean, family = binomial())
summary(fit3)
```
#Most significant factors are Age, HG, and ALT__SGPT_ 
```{r}
fit4 = glm(type ~ APC + FHIT + P15 + P73 + P14 + P16 + DAPK + RAR_ + RASSF1A + O6MGMT  + E_cadherin + methyl + age + gender + HG + AST__SGOT_ + ALT__SGPT_ + S_Albumin + bilirubin,data = dataclean, family = binomial())
summary(fit4)
```
# Not sure why when both fits are combined prob becomes zero or one, can't use stepwise for feature selection, go with most significant factors from both models. 
```{r}
fit5 = glm(type ~ APC + P15 + P14 + age + HG + ALT__SGPT_, data=dataclean, family = binomial )
summary(fit5)
```
#most significant from reduced model are p15, p14,age, and hg
```{r}
fit6 = glm(type ~ P15 + P14 + age + HG , data=dataclean, family = binomial )
summary(fit6)
```
```{r}
fit7 = glm(type ~ P15  + age + HG , data=dataclean, family = binomial )
summary(fit7)
```
#Final model type ~ P15  + age + HG

#SVM for predictive modeling 
```{r}
grid <- expand.grid(C =c(seq(0,5, by = 0.1)))
set.seed(1)
library(caret)
trctrl2 <- trainControl(method = "cv", number = 10, savePred=T)
svm1 <- train(type ~ P15  + age + HG,method = "svmLinear", trControl  = trctrl2,
data = dataclean, tuneGrid = grid)
svm1$results
```
#most efficient model predicts with 90% accuracy and tunign parameter 0.1.

#Run model on na.omit on hg only inorder to have more data points 
```{r}
dataclean2 = liver_data[!is.na(liver_data["HG"]),]
str(dataclean2)
```
```{r}
dataclean2[c] <- lapply(dataclean2[c], factor)
str(dataclean2)
```
```{r}
fit52 = glm(type ~ APC + P15 + P14 + age + HG + ALT__SGPT_, data=dataclean2, family = binomial )
summary(fit52)
```
```{r}
fit62 = glm(type ~ APC + P15 + P14 + age + HG, data=dataclean2, family = binomial )
summary(fit62)
```
#Final model for dataclean2 type ~ APC + P15 + P14  + age + HG

#SVM for predictive modeling 
```{r}
set.seed(1)
svm2 <- train(type ~ APC + P15 + P14 + age + HG,method = "svmLinear", trControl  = trctrl2,
data = dataclean2, tuneGrid = grid)
svm2$results
```

#SVM2 yields accuracy of 0.94667 with tuning parameter 0.2

#compare SVM2 to SVM1 on dataclean2
```{r}
set.seed(1)
svm1b <- train(type ~  P15  + age + HG,method = "svmLinear", trControl  = trctrl2,
data = dataclean2, tuneGrid = grid)
svm1b$results
```

#SVM1b yields accuracy of 0.93667 with tuning parameter 0,1

#In conclusion the most significant factors for determinig type are APC, P15, P14, age, and HG. SVM model used for predicting type has formula type ~ APC + P15 + P14  + age + HG with tuning parameter 0.2 and accuracy 0.94667